---
title: "Analysis of transmission type differences for cars of 1973."
output: pdf_document
---

```{r Loading libraries and data, echo = FALSE}
data(mtcars)
suppressMessages(library(ggplot2))
suppressMessages(library(GGally))
suppressMessages(library(dplyr))
```

```{r building table of p-values for am coefficient in different models, echo = FALSE}

```

```{r constructing models with significant am coefficients, echo = FALSE}
data(mtcars)
mtcars = mutate(mtcars, am = factor(am), vs = factor(vs))

fitAm = lm(mpg ~ am, mtcars)
fitAmQsec = lm(mpg ~ am + qsec , mtcars)
fitAmHp = lm(mpg ~ am + hp , mtcars)
fitAmVs = lm(mpg ~ am + vs , mtcars)
```

## Executive summary.
The fulfilled analysis revealed an insufficiency of data in the used dataset to 
perform causal analysis and discover if changing just transmission type causes
the significant changes of miles/gallon for automobiles of 1973. But results of the 
analysis also allow to believe that if exactly different transmission types cause
changes in miles/gallon (instead of some 3rd force causes changes in both of them) 
then manual transmission type has more miles/gallon than automatic one does.

## Analysis.
During exploratory analysis I plot 2 graphics and looked at the first rows of data.
```{r exploratory analysis, echo = FALSE}
data(mtcars)
gPairs = ggpairs(mtcars, axisLabels = "none",
    title = "So called, pairs graphic with correlation values for all variables of dataset.")

mtcars = mutate(mtcars, am = factor(am), vs = factor(vs), gear = factor(gear), carb = factor(carb), cyl = factor(cyl))
gMpgVsAm = ggplot(data = mtcars, aes(x = am, y = mpg)) + 
    geom_boxplot() +  
    labs(y = "miles/gallon") + 
    labs(title = "Miles/gallon dependency from transmission types.") +
    scale_x_discrete(name = "transmission type", breaks = c(0, 1), 
                     labels = c('automatic', 'manual'))


head(mtcars, 4)
```
From the boxplot (look at appendix) I can assume that manual transmission type must
have bigger amount of miles/gallon, but there are many other variables here, so 
may be there is some variable which causes both changes in transmission type and miles/gallon.
From the pairs graphic it is clear that transmission type ("am" on the plot) has strong correlation with many variables 
which also says for high probability of existence another force which causes changes in both 
transmission type and miles/gallon.

Because of the little quantity of observations only simple models (e.g. with 1-2 terms without intersection) have 
low p-values for all coefficients, which means that there coefficients are truly not zeros. I am interested only 
in models with am term (transmission type) included. Below there is the list of maximum of p-values 
among all coefficients for each model with 2 terms: am and the one specified in column name.

```{r looking for simple model which coefficients are all significant, echo=FALSE}
data(mtcars)
mtcars = mutate(mtcars, am = factor(am), vs = factor(vs), gear = factor(gear), carb = factor(carb), cyl = factor(cyl))

getPValuesFor2TermModels = function() {
    res = numeric()
    for (var in names(mtcars)) {
        if (var != "am" && var != "mpg") {
            formula = paste("mpg ~ am +", var)
            fit = lm(formula, data = mtcars)
            #saving maximum p-value of all coefficient for that model
            pRes = max(summary(fit)$coefficients[, "Pr(>|t|)"])
            names(pRes) = var
            res = c(res, round(pRes, 4))
        }
    }
    res
}

getPValuesFor2TermModels()
```
So, only the models with am and one term of qsec (1/4 mile time), hp (Gross horsepower), vs (V/S) have all coefficients 
which are significantly different from zeros (p-value < 0.05). All other models are uninterpretable 
because some of their coefficients probably equal to zero (We cannot reject this hypothesis).  
These 3 models and the model with only 1 term (am) I compared by ANOVA to find the one 
with the least sum of residual's squares (RSS column):

```{r ANOVA comparison, echo = FALSE}
fitAm = lm(mpg ~ am, mtcars)
fitAmQsec = lm(mpg ~ am + qsec , mtcars)
fitAmHp = lm(mpg ~ am + hp , mtcars)
fitAmVs = lm(mpg ~ am + factor(vs) , mtcars)
anova(fitAm, fitAmQsec, fitAmHp, fitAmVs)
```
So, the best model is the one with am and hp terms.  

Must to say that for all those 4 models the coefficient for am1 term is positive. This coefficient defines differences 
between manual (labeled as 1) and automatic (labeled as 0) transmission types given all other predictors 
(if they present) stay constant. So I can assume that if changes in transmission type 
can cause changes in miles/gallon values then for manual transmission type the value of miles/gallon is bigger 
than for automatic type. Below there are actual values of coefficients for each of those models 
with 95% confidence intervals:

```{r coefficient calculation, echo = FALSE}

printModel = function(model) {
    form = formula(model)
    paste(form[2], form[1], form[3])
}

getAmCoefficientAndConfidenceInterval = function(modelList) {
    res = data.frame(Formula = character(), CoefVal = numeric(), lowerConfInt = numeric(), upperConfInt = numeric())
    for (model in modelList) {
        coefStatist = summary(model)$coef
        confInt = 
            coefStatist['am1', 'Estimate'] + c(-1, 1) * qt( 0.975, model$df.residual) * coefStatist['am1', 'Std. Error']
        crntRes = data.frame(Formula = printModel(model), 
                            CoefVal = coefStatist['am1', 'Estimate'], 
                            lowerConfInt = confInt[1], 
                            upperConfInt = confInt[2])
        res = rbind(res, crntRes)
    }
    res
}

modelList = list(fitAm, fitAmQsec, fitAmHp, fitAmVs)
getAmCoefficientAndConfidenceInterval(modelList)
```
So, with the best model am + hp we can expect increasing in miles/gallon on 
`r getAmCoefficientAndConfidenceInterval(list(fitAmHp))$CoefVal` when we replace automatic transmission by manual one 
given the gross horsepower value is fixed. The uncertainty of this estimation for 95% confidence interval is 
(`r getAmCoefficientAndConfidenceInterval(list(fitAmHp))$lowerConfInt`; 
`r getAmCoefficientAndConfidenceInterval(list(fitAmHp))$upperConfInt`).

To perform some diagnostics I have placed residual plot and fitted value plot for the model mpg ~ am + hp 
in the appendix and below there are influence measures for mpg ~ am and mpg ~ am + hp models:
```{r fitted values plot, echo = FALSE}
# look for residual plot code in the appendix

gFittedValues = 
    ggplot() +
    geom_point(aes(x = fitAmHp$model$hp, y = fitAmHp$model$mpg, color = fitAmHp$model$am)) + 
    geom_line(aes(x = fitAmHp$model$hp, y = fitAmHp$fitted.values, color = fitAmHp$model$am)) +
    scale_color_manual(name = "Transmission\ntypes", breaks = c(0, 1), values = c("red", "green"),
            labels = c("automatic", "manual")) +
    labs(x = "Gross horsepower") +
    labs(y = "Miles/(US) gallon") +
    labs(title = "Estimated values of miles/gallon by model mpg ~ am + hp")


```

```{r influence measures, echo=FALSE}
summary(influence.measures(fitAm))
summary(influence.measures(fitAmHp))
```

I didn't exclude influential observations because there is too little data to reduce it further.

In the end I must to say that these models are not enough to conclude that changes in transmission type can 
cause change in miles/gallon. To proof such statement I need at least to fit the model with all possible terms,
get significant p-values for all coefficients, look at coefficient for am1 
(difference between manual transmission type and automatic one given all other predictors are constant)
and only then claim how exaclty transmission type affect the miles/gallon. But in the given dataset there is not enough data 
to get significant p-value for am1 coefficient (I cannot reject hypothesis that it is not equal to zero), 
but may be more data will not change this and then it means that transmission type doesn't affect miles/gallon at all.

\newpage 

## Appendix
### Exploratory analysis
```{r ,echo = FALSE}
gMpgVsAm
```

\newpage


```{r plot pairs graphic, echo=FALSE, fig.height=10, fig.width=10}
gPairs
```

\newpage

### Model analysis.
```{r fitted values, echo = FALSE, fig.height=4}
gFittedValues
```

```{r residuals, echo = FALSE, fig.height=4.5}
par(mfrow = c(2, 2))
plot(fitAmHp)
```
